---
title: "Enhanced Crime Analysis for Washington DC"
subtitle: "Spatial and Temporal Patterns"
author: 
  - name: "Troy Cheng\u00A0   Ziqiao Shan\u00A0  Minji Kang\u00A0 & \u00A0Sarah Krause"
    email: yc1317@georgetown.edu
    affiliation: Georgetown University
    corresponding: true
df-print: kable
title-block-banner: "#0a0e1a"
title-block-banner-color: "#4DB8FF"
execute:
  warning: false
date: 2025-03-25
date-modified: last-modified
format:
  html:
    embed-resources: true
    toc: true                 
    toc-title: "Contents"     
    toc-location: right       
    number-sections: true     
    smooth-scroll: true       
    css: trycstyle.css 
    code-overflow: wrap
include-in-header:
  text: |
    <link rel="shortcut icon" href="assets/gu.ico" type="image/x-icon">           
highlight-style: nord       
engine: knitr
---

# Enhanced Crime Analysis for Washington DC

## Load Packages and Data

```{r}
# Load required packages
# install.packages("leaflet.extras")
library(tidyverse)
library(sf)
library(ggmap)
library(leaflet)
library(leaflet.extras) # For heat maps and additional leaflet functionality
library(lubridate) # For date manipulation
library(scales)    # For better axis formatting
library(viridis)   # For color palettes
library(gridExtra) # For arranging multiple plots
library(DT)        # For interactive tables
library(plotly)    # For interactive plots
library(heatmaply) # For heat maps
```

```{r}
# Set the data directory path
data_dir <- "data/DC/"

# Load crime incidents data for multiple years
crime_2021 <- read_csv(paste0(data_dir, "crime_incidents/Crime_Incidents_in_2021.csv"))
crime_2022 <- read_csv(paste0(data_dir, "crime_incidents/Crime_Incidents_in_2022.csv"))
crime_2023 <- read_csv(paste0(data_dir, "crime_incidents/Crime_Incidents_in_2023.csv"))

# Load demographic data
demographic_data <- read_csv(paste0(data_dir, "acs5/ACS_5-Year_Demographic_Characteristics_DC.csv"))

# Load economic data
economic_data <- read_csv(paste0(data_dir, "acs5/ACS_5-Year_Economic_Characteristics_DC.csv"))

# Load housing data
housing_data <- read_csv(paste0(data_dir, "acs5/ACS_5-Year_Housing_Characteristics_DC_Census_Tract.csv"))
```

## Data Cleaning and Preparation

```{r}
# Clean and prepare crime data for 2021
crime_2021_clean <- crime_2021 %>%
  # Convert date columns to proper date format
  mutate(
    REPORT_DAT = as.Date(str_extract(REPORT_DAT, "\\d{4}/\\d{2}/\\d{2}")),
    START_DATE = as.Date(str_extract(START_DATE, "\\d{4}/\\d{2}/\\d{2}")),
    END_DATE = as.Date(str_extract(END_DATE, "\\d{4}/\\d{2}/\\d{2}"))
  ) %>%
  # Remove rows with missing coordinates
  filter(!is.na(LATITUDE) & !is.na(LONGITUDE)) %>%
  # Add year column
  mutate(YEAR = 2021)

# Clean and prepare crime data for 2022
crime_2022_clean <- crime_2022 %>%
  # Convert date columns to proper date format
  mutate(
    REPORT_DAT = as.Date(str_extract(REPORT_DAT, "\\d{4}/\\d{2}/\\d{2}")),
    START_DATE = as.Date(str_extract(START_DATE, "\\d{4}/\\d{2}/\\d{2}")),
    END_DATE = as.Date(str_extract(END_DATE, "\\d{4}/\\d{2}/\\d{2}"))
  ) %>%
  # Remove rows with missing coordinates
  filter(!is.na(LATITUDE) & !is.na(LONGITUDE)) %>%
  # Add year column
  mutate(YEAR = 2022)

# Clean and prepare crime data for 2023
crime_2023_clean <- crime_2023 %>%
  # Convert date columns to proper date format
  mutate(
    REPORT_DAT = as.Date(str_extract(REPORT_DAT, "\\d{4}/\\d{2}/\\d{2}")),
    START_DATE = as.Date(str_extract(START_DATE, "\\d{4}/\\d{2}/\\d{2}")),
    END_DATE = as.Date(str_extract(END_DATE, "\\d{4}/\\d{2}/\\d{2}"))
  ) %>%
  # Remove rows with missing coordinates
  filter(!is.na(LATITUDE) & !is.na(LONGITUDE)) %>%
  # Add year column
  mutate(YEAR = 2023)

# Combine all years into one dataset
all_crime <- bind_rows(crime_2021_clean, crime_2022_clean, crime_2023_clean)

# Extract time of day from START_DATE for time-based analysis
all_crime <- all_crime %>%
  mutate(
    hour_of_day = hour(as.POSIXct(str_extract(START_DATE, "\\d{4}/\\d{2}/\\d{2}"), format="%Y/%m/%d")),
    day_of_week = wday(REPORT_DAT, label = TRUE),
    month = month(REPORT_DAT, label = TRUE),
    season = case_when(
      month %in% c("Dec", "Jan", "Feb") ~ "Winter",
      month %in% c("Mar", "Apr", "May") ~ "Spring",
      month %in% c("Jun", "Jul", "Aug") ~ "Summer",
      month %in% c("Sep", "Oct", "Nov") ~ "Fall"
    ),
    time_of_day = case_when(
      SHIFT == "MIDNIGHT" ~ "Night (12AM-8AM)",
      SHIFT == "DAY" ~ "Day (8AM-4PM)",
      SHIFT == "EVENING" ~ "Evening (4PM-12AM)",
      TRUE ~ "Unknown"
    )
  )
```

## Multi-Year Crime Trend Analysis

```{r, fig.width=10, fig.height=6}
# Count incidents by year
yearly_counts <- all_crime %>%
  count(YEAR) %>%
  mutate(YEAR = as.factor(YEAR))

# Create a bar chart of incidents by year
ggplot(yearly_counts, aes(x = YEAR, y = n, fill = YEAR)) +
  geom_col() +
  geom_text(aes(label = n), vjust = -0.5) +
  labs(title = "Total Crime Incidents by Year (2021-2023)",
       x = "Year",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_viridis_d() +
  theme(legend.position = "none")

# Count incidents by year and offense type
yearly_offense_counts <- all_crime %>%
  count(YEAR, OFFENSE) %>%
  group_by(YEAR) %>%
  top_n(10, n) %>%
  ungroup() %>%
  mutate(YEAR = as.factor(YEAR))

# Create a bar chart of top offenses by year
ggplot(yearly_offense_counts, aes(x = reorder(OFFENSE, n), y = n, fill = YEAR)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(title = "Top 10 Crime Types by Year (2021-2023)",
       x = "Crime Type",
       y = "Number of Incidents",
       fill = "Year") +
  theme_minimal() +
  scale_fill_viridis_d()

# Calculate percent change from 2021 to 2023
offense_trend <- all_crime %>%
  count(YEAR, OFFENSE) %>%
  group_by(OFFENSE) %>%
  filter(n() == 3) %>%  # Only include offenses present in all 3 years
  mutate(pct_change = (n[YEAR == 2023] - n[YEAR == 2021]) / n[YEAR == 2021] * 100) %>%
  filter(YEAR == 2023) %>%  # Just keep the 2023 row for each offense
  ungroup() %>%
  top_n(10, abs(pct_change))  # Top 10 by absolute percent change

# Create a bar chart of percent change
ggplot(offense_trend, aes(x = reorder(OFFENSE, pct_change), y = pct_change, fill = pct_change > 0)) +
  geom_col() +
  coord_flip() +
  labs(title = "Percent Change in Crime Types (2021 to 2023)",
       x = "Crime Type",
       y = "Percent Change",
       fill = "Increase") +
  theme_minimal() +
  scale_fill_manual(values = c("darkred", "darkgreen")) +
  geom_text(aes(label = sprintf("%.1f%%", pct_change)), hjust = ifelse(offense_trend$pct_change > 0, -0.1, 1.1)) +
  theme(legend.position = "none")
```

## Temporal Analysis: Crime by Time of Day

```{r, fig.width=10, fig.height=8}
# Analyze crime by shift (time of day)
shift_counts <- all_crime %>%
  count(SHIFT) %>%
  filter(!is.na(SHIFT))

# Create a pie chart of incidents by shift
ggplot(shift_counts, aes(x = "", y = n, fill = SHIFT)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(title = "Crime Incidents by Shift (2021-2023)",
       fill = "Shift") +
  theme_void() +
  scale_fill_viridis_d() +
  geom_text(aes(label = paste0(round(n/sum(n)*100, 1), "%")), 
            position = position_stack(vjust = 0.5))

# Analyze crime by day of week
day_counts <- all_crime %>%
  count(day_of_week) %>%
  filter(!is.na(day_of_week))

# Create a bar chart of incidents by day of week
ggplot(day_counts, aes(x = day_of_week, y = n, fill = day_of_week)) +
  geom_col() +
  labs(title = "Crime Incidents by Day of Week (2021-2023)",
       x = "Day of Week",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_viridis_d() +
  theme(legend.position = "none")

# Create a heatmap of crime by day of week and shift
day_shift_counts <- all_crime %>%
  count(day_of_week, SHIFT) %>%
  filter(!is.na(day_of_week) & !is.na(SHIFT))

# Reshape data for heatmap
day_shift_matrix <- day_shift_counts %>%
  pivot_wider(names_from = SHIFT, values_from = n) %>%
  column_to_rownames(var = "day_of_week")

# Create a heatmap
heatmap(as.matrix(day_shift_matrix), 
        Colv = NA, Rowv = NA, 
        scale = "column", 
        col = viridis(100),
        main = "Crime Incidents by Day of Week and Shift",
        xlab = "Shift", 
        ylab = "Day of Week")

# Analyze crime by month and season
month_counts <- all_crime %>%
  count(month) %>%
  filter(!is.na(month))

# Create a bar chart of incidents by month
ggplot(month_counts, aes(x = month, y = n, fill = month)) +
  geom_col() +
  labs(title = "Crime Incidents by Month (2021-2023)",
       x = "Month",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_viridis_d() +
  theme(legend.position = "none")

# Analyze crime by season
season_counts <- all_crime %>%
  count(season) %>%
  filter(!is.na(season))

# Create a bar chart of incidents by season
ggplot(season_counts, aes(x = season, y = n, fill = season)) +
  geom_col() +
  labs(title = "Crime Incidents by Season (2021-2023)",
       x = "Season",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_viridis_d() +
  theme(legend.position = "none")
```

## Spatial Analysis: Crime Heat Maps

```{r, fig.width=12, fig.height=10}
# Create a leaflet heat map of all crime incidents
leaflet(all_crime) %>%
  addTiles() %>%
  addHeatmap(
    lng = ~LONGITUDE, 
    lat = ~LATITUDE, 
    intensity = 0.5,
    radius = 10, 
    blur = 15,
    max = 0.05,
    gradient = "RdYlBu"
  ) %>%
  addLegend("bottomright", 
            colors = c("blue", "yellow", "red"), 
            labels = c("Low", "Medium", "High"), 
            title = "Crime Density")

# Create separate heat maps for different crime types
top_crimes <- all_crime %>%
  count(OFFENSE, sort = TRUE) %>%
  head(5) %>%
  pull(OFFENSE)

# Filter data for top crime types
top_crime_data <- all_crime %>%
  filter(OFFENSE %in% top_crimes)

# Create a leaflet map with layers for each crime type
leaflet() %>%
  addTiles() %>%
  addLayersControl(
    overlayGroups = top_crimes,
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  # Add heat maps for each crime type
  addHeatmap(
    data = filter(top_crime_data, OFFENSE == top_crimes[1]),
    lng = ~LONGITUDE, 
    lat = ~LATITUDE, 
    intensity = 0.5,
    radius = 10, 
    blur = 15,
    max = 0.05,
    group = top_crimes[1]
  ) %>%
  addHeatmap(
    data = filter(top_crime_data, OFFENSE == top_crimes[2]),
    lng = ~LONGITUDE, 
    lat = ~LATITUDE, 
    intensity = 0.5,
    radius = 10, 
    blur = 15,
    max = 0.05,
    group = top_crimes[2]
  ) %>%
  addHeatmap(
    data = filter(top_crime_data, OFFENSE == top_crimes[3]),
    lng = ~LONGITUDE, 
    lat = ~LATITUDE, 
    intensity = 0.5,
    radius = 10, 
    blur = 15,
    max = 0.05,
    group = top_crimes[3]
  ) %>%
  addHeatmap(
    data = filter(top_crime_data, OFFENSE == top_crimes[4]),
    lng = ~LONGITUDE, 
    lat = ~LATITUDE, 
    intensity = 0.5,
    radius = 10, 
    blur = 15,
    max = 0.05,
    group = top_crimes[4]
  ) %>%
  addHeatmap(
    data = filter(top_crime_data, OFFENSE == top_crimes[5]),
    lng = ~LONGITUDE, 
    lat = ~LATITUDE, 
    intensity = 0.5,
    radius = 10, 
    blur = 15,
    max = 0.05,
    group = top_crimes[5]
  )
```

## Crime and Method Analysis

```{r, fig.width=10, fig.height=6}
# Analyze crime methods by year
method_by_year <- all_crime %>%
  count(YEAR, METHOD) %>%
  group_by(YEAR) %>%
  top_n(5, n) %>%
  ungroup() %>%
  mutate(YEAR = as.factor(YEAR))

# Create a bar chart of top methods by year
ggplot(method_by_year, aes(x = reorder(METHOD, n), y = n, fill = YEAR)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(title = "Top 5 Crime Methods by Year (2021-2023)",
       x = "Method",
       y = "Number of Incidents",
       fill = "Year") +
  theme_minimal() +
  scale_fill_viridis_d()

# Analyze relationship between crime type and method
crime_method <- all_crime %>%
  count(OFFENSE, METHOD) %>%
  group_by(OFFENSE) %>%
  top_n(5, n) %>%
  ungroup() %>%
  filter(OFFENSE %in% top_crimes)

# Create a stacked bar chart of methods by crime type
ggplot(crime_method, aes(x = OFFENSE, y = n, fill = METHOD)) +
  geom_col(position = "fill") +
  labs(title = "Crime Methods by Offense Type (2021-2023)",
       x = "Offense Type",
       y = "Proportion",
       fill = "Method") +
  theme_minimal() +
  scale_fill_viridis_d() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Ward-Level Analysis

```{r, fig.width=10, fig.height=6}
# Analyze crime by ward and year
ward_by_year <- all_crime %>%
  filter(!is.na(WARD) & WARD != "") %>%
  count(YEAR, WARD) %>%
  mutate(YEAR = as.factor(YEAR))

# Create a bar chart of incidents by ward and year
ggplot(ward_by_year, aes(x = WARD, y = n, fill = YEAR)) +
  geom_col(position = "dodge") +
  labs(title = "Crime Incidents by Ward and Year (2021-2023)",
       x = "Ward",
       y = "Number of Incidents",
       fill = "Year") +
  theme_minimal() +
  scale_fill_viridis_d()

# Analyze top crime types by ward
ward_crime_types <- all_crime %>%
  filter(!is.na(WARD) & WARD != "") %>%
  count(WARD, OFFENSE) %>%
  group_by(WARD) %>%
  top_n(3, n) %>%
  ungroup()

# Create a faceted bar chart of top crime types by ward
ggplot(ward_crime_types, aes(x = reorder(OFFENSE, n), y = n, fill = OFFENSE)) +
  geom_col() +
  facet_wrap(~WARD, scales = "free_y") +
  coord_flip() +
  labs(title = "Top 3 Crime Types by Ward (2021-2023)",
       x = "Crime Type",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_viridis_d() +
  theme(legend.position = "none")
```

## Conclusion

This enhanced analysis of crime data in Washington DC from 2021 to 2023 reveals several important patterns:

1. **Temporal Patterns**: Crime incidents show distinct patterns by time of day, day of week, and season, with certain periods experiencing higher rates of specific crimes.

2. **Spatial Distribution**: Heat maps highlight crime hotspots across the city, with different crime types showing unique spatial patterns.

3. **Crime Trends**: The analysis of multi-year data shows changing patterns in crime types and methods over time, with some offenses increasing while others decrease.

4. **Ward-Level Variations**: Different wards experience different types and frequencies of crime, suggesting the need for targeted prevention strategies.

These insights can inform policy decisions, resource allocation, and crime prevention strategies in Washington DC. Further analysis could explore the relationships between crime patterns and socioeconomic factors, as well as the effectiveness of various intervention strategies.