---
title: "Data Science for Public Policy"
subtitle: "Final Project"
author: 
  - name: "Troy Cheng\u00A0   Ziqiao Shan\u00A0  Minji Kang\u00A0 & \u00A0Sarah Krause"
    email: yc1317@georgetown.edu
    affiliation: Georgetown University
    corresponding: true
df-print: kable
title-block-banner: "#0a0e1a"
title-block-banner-color: "#4DB8FF"
execute:
  warning: false
date: 2025-03-25
date-modified: last-modified
format:
  html:
    embed-resources: true
    toc: true                 
    toc-title: "Contents"     
    toc-location: right       
    number-sections: true     
    smooth-scroll: true       
    css: trycstyle.css 
    code-overflow: wrap
include-in-header:
  text: |
    <link rel="shortcut icon" href="assets/gu.ico" type="image/x-icon">           
highlight-style: nord       
engine: knitr
---

# Exploratory Data Analysis of DC Crime Incidents (2008-2025)

## Introduction

This analysis explores crime incident data in Washington DC from 2008 to 2025, focusing on spatial and temporal patterns. We'll examine trends in crime types, geographic distribution, and changes over time to provide insights into crime patterns across the city. The analysis uses the comprehensive crime incidents dataset and integrates spatial data to visualize crime patterns across different areas of DC.

## Load Packages

```{r}
# Load required packages
# install.packages(c("leaflet", "scales", "viridis", "leaflet.extras", "DT", "plotly", "patchwork"))

library(tidyverse)
library(sf)
library(ggmap)
library(leaflet)
library(leaflet.extras) # For heat maps and additional leaflet functionality
library(lubridate)      # For date manipulation
library(scales)         # For better axis formatting
library(viridis)        # For color palettes
library(gridExtra)      # For arranging multiple plots
library(DT)             # For interactive tables
library(plotly)         # For interactive plots
library(patchwork)      # For combining plots
```

## Load and Prepare Crime Data

```{r}
# Set the data directory path
data_dir <- "data/DC/crime_incidents/"

# Load the comprehensive crime data (2008-2025)
all_crime <- read_csv(paste0(data_dir, "Crime_Incidents_in_2008 - 2025.csv"))

# Clean the data
all_crime_clean <- all_crime %>%
  # Convert date columns to proper date format
  mutate(
    REPORT_DAT = as.Date(str_extract(REPORT_DAT, "\\d{4}/\\d{2}/\\d{2}")),
    START_DATE = as.Date(str_extract(START_DATE, "\\d{4}/\\d{2}/\\d{2}")),
    END_DATE = as.Date(str_extract(END_DATE, "\\d{4}/\\d{2}/\\d{2}"))
  ) %>%
  # Extract year from report date
  mutate(YEAR = year(REPORT_DAT)) %>%
  # Remove rows with missing coordinates
  filter(!is.na(LATITUDE) & !is.na(LONGITUDE))

# Display the structure of the dataset
glimpse(all_crime_clean)
```

## Basic Crime Statistics

```{r}
# Count of incidents by year
all_crime_clean %>%
  count(YEAR) %>%
  arrange(YEAR) %>%
  knitr::kable(caption = "Number of Crime Incidents by Year (2008-2025)")

# Visualize crime trends over years
all_crime_clean %>%
  count(YEAR) %>%
  ggplot(aes(x = YEAR, y = n)) +
  geom_line(color = "#4DB8FF", size = 1) +
  geom_point(color = "#0a0e1a", size = 2) +
  labs(title = "Crime Incidents Trend (2008-2025)",
       x = "Year",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(2008, 2025, 2)) +
  scale_y_continuous(labels = comma)

# Top crime types across all years
all_crime_clean %>%
  count(OFFENSE, sort = TRUE) %>%
  head(10) %>%
  ggplot(aes(x = reorder(OFFENSE, n), y = n, fill = n)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top 10 Crime Types (2008-2025)",
       x = "Crime Type",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_viridis_c() +
  scale_y_continuous(labels = comma) +
  theme(legend.position = "none")

# Crime methods across all years
all_crime_clean %>%
  count(METHOD, sort = TRUE) %>%
  head(10) %>%
  ggplot(aes(x = reorder(METHOD, n), y = n, fill = n)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top 10 Crime Methods (2008-2025)",
       x = "Method",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_viridis_c() +
  scale_y_continuous(labels = comma) +
  theme(legend.position = "none")

# Crime by ward
all_crime_clean %>%
  filter(!is.na(WARD) & WARD != "") %>%
  count(WARD, sort = TRUE) %>%
  ggplot(aes(x = reorder(WARD, n), y = n, fill = n)) +
  geom_col() +
  labs(title = "Crime Incidents by Ward (2008-2025)",
       x = "Ward",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_viridis_c() +
  scale_y_continuous(labels = comma) +
  theme(legend.position = "none")

# Crime by shift
all_crime_clean %>%
  count(SHIFT, sort = TRUE) %>%
  ggplot(aes(x = reorder(SHIFT, n), y = n, fill = SHIFT)) +
  geom_col() +
  labs(title = "Crime Incidents by Shift (2008-2025)",
       x = "Shift",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_viridis_d() +
  scale_y_continuous(labels = comma) +
  theme(legend.position = "none")
```

## Temporal Analysis

```{r, fig.width=10, fig.height=8}
# Create a subset of recent years for more detailed analysis
recent_crime <- all_crime_clean %>%
  filter(YEAR >= 2018)

# Crime by month across recent years
p1 <- recent_crime %>%
  mutate(month = month(REPORT_DAT, label = TRUE)) %>%
  count(YEAR, month) %>%
  ggplot(aes(x = month, y = n, color = factor(YEAR), group = YEAR)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Crime Incidents by Month (2018-2025)",
       x = "Month",
       y = "Number of Incidents",
       color = "Year") +
  theme_minimal() +
  scale_color_viridis_d() +
  scale_y_continuous(labels = comma)

# Monthly patterns across all years
p2 <- all_crime_clean %>%
  mutate(month = month(REPORT_DAT, label = TRUE)) %>%
  count(month) %>%
  ggplot(aes(x = month, y = n, fill = n)) +
  geom_col() +
  labs(title = "Average Monthly Crime Pattern (2008-2025)",
       x = "Month",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_viridis_c() +
  scale_y_continuous(labels = comma) +
  theme(legend.position = "none")

# Combine the plots
p1 / p2

# Crime by day of week
p3 <- all_crime_clean %>%
  mutate(day_of_week = wday(REPORT_DAT, label = TRUE)) %>%
  count(day_of_week) %>%
  ggplot(aes(x = day_of_week, y = n, fill = day_of_week)) +
  geom_col() +
  labs(title = "Crime Incidents by Day of Week (2008-2025)",
       x = "Day of Week",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_viridis_d() +
  scale_y_continuous(labels = comma) +
  theme(legend.position = "none")

# Crime by time of day (shift)
p4 <- all_crime_clean %>%
  count(SHIFT) %>%
  ggplot(aes(x = SHIFT, y = n, fill = SHIFT)) +
  geom_col() +
  labs(title = "Crime Incidents by Time of Day (2008-2025)",
       x = "Shift",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_viridis_d() +
  scale_y_continuous(labels = comma) +
  theme(legend.position = "none")

# Combine day of week and shift plots
p3 + p4

# Heatmap of crime by year and month
all_crime_clean %>%
  mutate(month = month(REPORT_DAT, label = TRUE)) %>%
  count(YEAR, month) %>%
  ggplot(aes(x = month, y = factor(YEAR), fill = n)) +
  geom_tile() +
  labs(title = "Crime Incidents Heatmap by Year and Month",
       x = "Month",
       y = "Year",
       fill = "Number of Incidents") +
  theme_minimal() +
  scale_fill_viridis_c() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Seasonal patterns by crime type (top 5)
all_crime_clean %>%
  mutate(month = month(REPORT_DAT, label = TRUE)) %>%
  count(OFFENSE, month) %>%
  group_by(OFFENSE) %>%
  mutate(total = sum(n)) %>%
  ungroup() %>%
  arrange(desc(total)) %>%
  filter(OFFENSE %in% unique(OFFENSE)[1:5]) %>%
  ggplot(aes(x = month, y = n, color = OFFENSE, group = OFFENSE)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Seasonal Patterns for Top 5 Crime Types",
       x = "Month",
       y = "Number of Incidents",
       color = "Crime Type") +
  theme_minimal() +
  scale_y_continuous(labels = comma)
```

## Crime Type Analysis

```{r, fig.width=10, fig.height=10}
# Create a subset of recent years for more detailed analysis
recent_years <- c(2020, 2021, 2022, 2023, 2024, 2025)
recent_crime <- all_crime_clean %>%
  filter(YEAR %in% recent_years)

# Top crime types by recent years
recent_crime %>%
  count(YEAR, OFFENSE) %>%
  group_by(YEAR) %>%
  top_n(5, n) %>%
  ungroup() %>%
  ggplot(aes(x = reorder(OFFENSE, n), y = n, fill = factor(YEAR))) +
  geom_col() +
  facet_wrap(~YEAR, scales = "free_y", ncol = 2) +
  coord_flip() +
  labs(title = "Top 5 Crime Types by Year (2020-2025)",
       x = "Crime Type",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_viridis_d() +
  scale_y_continuous(labels = comma) +
  theme(legend.position = "none")

# Trend of top 5 crime types over all years
top_5_crimes <- all_crime_clean %>%
  count(OFFENSE) %>%
  top_n(5, n) %>%
  pull(OFFENSE)

all_crime_clean %>%
  filter(OFFENSE %in% top_5_crimes) %>%
  count(YEAR, OFFENSE) %>%
  ggplot(aes(x = YEAR, y = n, color = OFFENSE, group = OFFENSE)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Trend of Top 5 Crime Types (2008-2025)",
       x = "Year",
       y = "Number of Incidents",
       color = "Crime Type") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(2008, 2025, 2)) +
  scale_y_continuous(labels = comma)

# Crime methods by recent years
recent_crime %>%
  filter(METHOD != "OTHERS") %>%  # Exclude "OTHERS" to focus on specific methods
  count(YEAR, METHOD) %>%
  group_by(YEAR) %>%
  top_n(5, n) %>%
  ungroup() %>%
  ggplot(aes(x = reorder(METHOD, n), y = n, fill = factor(YEAR))) +
  geom_col() +
  facet_wrap(~YEAR, scales = "free_y", ncol = 2) +
  coord_flip() +
  labs(title = "Top 5 Crime Methods by Year (2020-2025)",
       x = "Method",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_viridis_d() +
  scale_y_continuous(labels = comma) +
  theme(legend.position = "none")

# Trend of violent vs. property crimes
all_crime_clean <- all_crime_clean %>%
  mutate(crime_category = case_when(
    OFFENSE %in% c("HOMICIDE", "ASSAULT W/DANGEROUS WEAPON", "SEX ABUSE", "ROBBERY") ~ "Violent Crime",
    OFFENSE %in% c("THEFT/OTHER", "THEFT F/AUTO", "BURGLARY", "MOTOR VEHICLE THEFT", "ARSON") ~ "Property Crime",
    TRUE ~ "Other Crime"
  ))

all_crime_clean %>%
  count(YEAR, crime_category) %>%
  ggplot(aes(x = YEAR, y = n, color = crime_category, group = crime_category)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Trend of Crime Categories (2008-2025)",
       x = "Year",
       y = "Number of Incidents",
       color = "Crime Category") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(2008, 2025, 2)) +
  scale_y_continuous(labels = comma) +
  scale_color_viridis_d()

# Proportion of crime categories by year
all_crime_clean %>%
  count(YEAR, crime_category) %>%
  group_by(YEAR) %>%
  mutate(proportion = n / sum(n)) %>%
  ggplot(aes(x = YEAR, y = proportion, fill = crime_category)) +
  geom_area() +
  labs(title = "Proportion of Crime Categories by Year (2008-2025)",
       x = "Year",
       y = "Proportion",
       fill = "Crime Category") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(2008, 2025, 2)) +
  scale_y_continuous(labels = percent) +
  scale_fill_viridis_d()
```

## Spatial Analysis

```{r, fig.width=10, fig.height=10}
# Load the shapefile for DC crime incidents from 2025
dc_crime_sp <- st_read("data/DC/crime_incidents/shp/Crime_Incidents_in_2025/Crime_Incidents_in_2025.shp", quiet = TRUE)

# Load police sectors shapefile
police_sectors_sp <- st_read("data/DC/polica_sector_division/Police_Sectors/Police_Sectors.shp", quiet = TRUE)

# Transform police sectors to WGS84 (EPSG:4326) to match crime data
police_sectors_sp <- st_transform(police_sectors_sp, crs = 4326)

# Load census blocks shapefile
census_blocks_sp <- st_read("data/DC/2020_census_blockLevel/Census_Blocks_in_2020/Census_Blocks_in_2020.shp", quiet = TRUE)

# Convert crime data to sf object for spatial analysis
crime_sf <- all_crime_clean %>%
  filter(YEAR == 2025) %>%  # Focus on 2025 for spatial analysis
  st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4326)

# Create a basic map of DC with crime incidents
leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = police_sectors_sp, 
              fillColor = "#4DB8FF",
              fillOpacity = 0.2,
              color = "#0a0e1a",
              weight = 1,
              label = ~NAME) %>%
  addCircleMarkers(data = crime_sf,
                  radius = 2,
                  color = "#FF4D4D",
                  fillOpacity = 0.7,
                  stroke = FALSE,
                  popup = ~paste("Type: ", OFFENSE, "<br>",
                                "Date: ", REPORT_DAT, "<br>",
                                "Method: ", METHOD)) %>%
  addLegend(position = "bottomright",
            colors = c("#4DB8FF", "#FF4D4D"),
            labels = c("Police Sectors", "Crime Incidents"),
            title = "DC Crime Map (2025)")

# Create a heatmap of crime incidents
leaflet() %>%
  addProviderTiles(providers$CartoDB.DarkMatter) %>%
  addHeatmap(data = crime_sf,
             radius = 10,
             blur = 15,
             max = 0.1,
             gradient = "RdYlBu") %>%
  addLegend(position = "bottomright",
            colors = colorRampPalette(c("blue", "yellow", "red"))(3),
            labels = c("Low", "Medium", "High"),
            title = "Crime Density (2025)")

# Perform spatial join with police sectors
crime_sectors_joined <- st_join(crime_sf, police_sectors_sp, join = st_intersects) %>%
  st_drop_geometry()

# Aggregate crime by police sector
sector_crime_counts <- crime_sectors_joined %>%
  group_by(SECTOR) %>%
  summarize(crime_count = n()) %>%
  arrange(desc(crime_count))

# Display top sectors by crime count
head(sector_crime_counts, 10) %>%
  knitr::kable(caption = "Top 10 Police Sectors by Crime Count (2025)")

# Join crime counts back to police sectors for choropleth map
sector_crime_map <- left_join(police_sectors_sp, sector_crime_counts, by = "SECTOR")

# Create choropleth map of crime by police sector
ggplot(sector_crime_map) +
  geom_sf(aes(fill = crime_count)) +
  scale_fill_viridis_c(option = "plasma", name = "Crime Count") +
  labs(title = "Crime Incidents by Police Sector (2025)") +
  theme_minimal()

# Analyze crime by type and location
crime_type_location <- crime_sf %>%
  group_by(OFFENSE) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>%
  head(5)

# Create maps for top 5 crime types
top_5_crime_types <- crime_type_location$OFFENSE

for (crime_type in top_5_crime_types) {
  crime_subset <- crime_sf %>% filter(OFFENSE == crime_type)
  
  print(
    leaflet() %>%
      addProviderTiles(providers$CartoDB.Positron) %>%
      addCircleMarkers(data = crime_subset,
                      radius = 3,
                      color = "#FF4D4D",
                      fillOpacity = 0.7,
                      stroke = FALSE) %>%
      addLegend(position = "bottomright",
                colors = "#FF4D4D",
                labels = crime_type,
                title = paste("Crime Type:", crime_type))
  )
}

# Analyze crime patterns by time of day and location
crime_time_location <- crime_sf %>%
  group_by(SHIFT) %>%
  summarize(count = n())

# Create maps for each time shift
for (shift in unique(crime_sf$SHIFT)) {
  crime_subset <- crime_sf %>% filter(SHIFT == shift)
  
  print(
    leaflet() %>%
      addProviderTiles(providers$CartoDB.Positron) %>%
      addHeatmap(data = crime_subset,
                radius = 10,
                blur = 15,
                max = 0.1,
                gradient = "RdYlBu") %>%
      addLegend(position = "bottomright",
                colors = colorRampPalette(c("blue", "yellow", "red"))(3),
                labels = c("Low", "Medium", "High"),
                title = paste("Crime Density -", shift))
  )
}
# Create a basic map of crime incidents for 2023
try({
  crime_2023_clean <- all_crime_clean %>% filter(YEAR == 2023)
  
  leaflet(crime_2023_clean) %>%
    addTiles() %>%  # Add default OpenStreetMap tiles
    addCircleMarkers(
      ~LONGITUDE, ~LATITUDE,
      radius = 3,
      color = "red",
      stroke = FALSE,
      fillOpacity = 0.5,
      popup = ~paste("Offense:", OFFENSE, "<br>Date:", REPORT_DAT, "<br>Block:", BLOCK)
    ) %>%
    addLegend("bottomright", 
              colors = "red",
              labels = "Crime Incidents (2023)",
              opacity = 0.7)
  
  # Create a heat map of crime incidents for 2023
  leaflet(crime_2023_clean) %>%
    addTiles() %>%  # Add default OpenStreetMap tiles
    addHeatmap(
      lng = ~LONGITUDE, 
      lat = ~LATITUDE,
      intensity = 0.5,
      radius = 10,
      blur = 15
    ) %>%
    addLegend("bottomright", 
              colors = c("yellow", "red"),
              labels = c("Low Density", "High Density"),
              title = "Crime Density (2023)",
              opacity = 0.7)
}, silent = TRUE)
```

## Load and Analyze Shapefile Data

```{r}
# Load shapefile data for 2025
crime_2025_sf <- st_read(paste0(data_dir, "shp/Crime_Incidents_in_2025/Crime_Incidents_in_2025.shp"))

# Display the structure of the shapefile data
glimpse(crime_2025_sf)

# Display the first few rows of the shapefile data
head(crime_2025_sf) %>% knitr::kable()

# Plot the spatial distribution of crime incidents from the shapefile
plot(st_geometry(crime_2025_sf), main = "Spatial Distribution of Crime Incidents (2025)")
```

## Combine CSV and Shapefile Data

```{r}
# Load the combined CSV data (2008-2025)
try({
  all_years_csv <- read_csv(paste0(data_dir, "Crime_Incidents_in_2008 - 2025.csv"))
  
  # Clean the combined CSV data
  all_years_clean <- all_years_csv %>%
    # Convert date columns to proper date format
    mutate(
      REPORT_DAT = as.Date(str_extract(REPORT_DAT, "\\d{4}/\\d{2}/\\d{2}")),
      START_DATE = as.Date(str_extract(START_DATE, "\\d{4}/\\d{2}/\\d{2}")),
      END_DATE = as.Date(str_extract(END_DATE, "\\d{4}/\\d{2}/\\d{2}"))
    ) %>%
    # Remove rows with missing coordinates
    filter(!is.na(LATITUDE) & !is.na(LONGITUDE))
  
  # Convert CSV data to sf object
  all_years_sf <- all_years_clean %>%
    st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4326)
  
  # Combine with shapefile data
  combined_sf <- rbind(all_years_sf, crime_2025_sf)
  
  # Plot the combined data
  plot(st_geometry(combined_sf), main = "Combined Crime Incidents (2008-2025)")
}, silent = TRUE)
```

## Crime Trends Over Time

```{r, fig.width=10, fig.height=6}
# Extract year from report date for all crime data
all_crime <- all_crime %>%
  mutate(year = year(REPORT_DAT))

# Count incidents by year
yearly_counts <- all_crime %>%
  count(year)

# Plot crime trends over time
ggplot(yearly_counts, aes(x = year, y = n)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "blue", size = 3) +
  labs(title = "Crime Incidents by Year",
       x = "Year",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_x_continuous(breaks = unique(yearly_counts$year))

# Count incidents by year and offense type (top 5)
top_offenses <- all_crime %>%
  count(OFFENSE) %>%
  top_n(5, n) %>%
  pull(OFFENSE)

offense_trends <- all_crime %>%
  filter(OFFENSE %in% top_offenses) %>%
  count(year, OFFENSE)

# Plot trends for top offense types
ggplot(offense_trends, aes(x = year, y = n, color = OFFENSE, group = OFFENSE)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Trends for Top 5 Crime Types",
       x = "Year",
       y = "Number of Incidents",
       color = "Offense Type") +
  theme_minimal() +
  scale_x_continuous(breaks = unique(offense_trends$year)) +
  scale_color_viridis_d()
```

## Ward-Level Analysis

```{r, fig.width=10, fig.height=8}
# # Crime incidents by ward and year
# ward_year_counts <- all_crime %>%
#   filter(!is.na(WARD) & WARD != "") %>%
#   count(WARD, YEAR)
# 
# # Plot crime by ward and year
# ggplot(ward_year_counts, aes(x = WARD, y = n, fill = factor(YEAR))) +
#   geom_col(position = "dodge") +
#   labs(title = "Crime Incidents by Ward and Year",
#        x = "Ward",
#        y = "Number of Incidents",
#        fill = "Year") +
#   theme_minimal() +
#   scale_fill_viridis_d()
# 
# # Analyze top crime types by ward
# ward_crime_types <- all_crime %>%
#   filter(!is.na(WARD) & WARD != "") %>%
#   count(WARD, OFFENSE) %>%
#   group_by(WARD) %>%
#   top_n(3, n) %>%
#   ungroup()
# 
# # Create a faceted bar chart of top crime types by ward
# ggplot(ward_crime_types, aes(x = reorder(OFFENSE, n), y = n, fill = OFFENSE)) +
#   geom_col() +
#   facet_wrap(~WARD, scales = "free_y") +
#   coord_flip() +
#   labs(title = "Top 3 Crime Types by Ward (2021-2023)",
#        x = "Crime Type",
#        y = "Number of Incidents") +
#   theme_minimal() +
#   scale_fill_viridis_d() +
#   theme(legend.position = "none")
```

# Conclusion

This exploratory data analysis of crime incidents in Washington DC from 2008 to 2025 reveals several important patterns:

1.  **Temporal Patterns**: Crime incidents show distinct patterns by time of day, day of week, and season, with certain periods experiencing higher rates of specific crimes.

2.  **Spatial Distribution**: Heat maps highlight crime hotspots across the city, with different crime types showing unique spatial patterns.

3.  **Crime Trends**: The analysis of multi-year data shows changing patterns in crime types and methods over time, with some offenses increasing while others decrease.

4.  **Ward-Level Variations**: Different wards experience different types and frequencies of crime, suggesting the need for targeted prevention strategies.

Further analysis could explore the relationship between crime patterns and socioeconomic factors, as well as the impact of policy interventions on crime rates in specific areas.

\`\`\`

{=html}

```{=html}
<script>
document.addEventListener("DOMContentLoaded", function() {
    const toc = document.getElementById("TOC");
    if (toc) {
        const sourceLink = document.createElement("div");
        sourceLink.innerHTML = `
            <div class="toc-source">
                <a href="https://github.com/troy-yu-cheng/final-project" 
                   target="_blank" 
                   class="github-button">
                   <svg xmlns="http://www.w3.org/2000/svg" 
                        viewBox="0 0 24 24" 
                        width="16" 
                        height="16" 
                        fill="currentColor"
                        style="vertical-align: middle; margin-right: 5px;">
                     <path d="M12 0C5.373 0 0 5.373 0 12c0 5.303 3.438 9.8 8.207 11.387.6.113.82-.26.82-.577v-2.157c-3.338.726-4.033-1.416-4.033-1.416-.546-1.386-1.332-1.756-1.332-1.756-1.09-.745.083-.73.083-.73 1.205.084 1.84 1.237 1.84 1.237 1.07 1.832 2.807 1.303 3.492.996.108-.774.418-1.303.76-1.602-2.665-.3-5.466-1.332-5.466-5.93 0-1.311.468-2.382 1.237-3.222-.124-.302-.536-1.52.118-3.163 0 0 1.008-.322 3.3 1.23a11.516 11.516 0 0 1 3.002-.403 11.486 11.486 0 0 1 3.002.403c2.292-1.552 3.3-1.23 3.3-1.23.654 1.644.242 2.861.118 3.163.77.84 1.236 1.911 1.236 3.222 0 4.61-2.807 5.627-5.48 5.922.43.372.812 1.103.812 2.222v3.293c0 .321.218.694.825.576C20.565 21.796 24 17.3 24 12 24 5.373 18.627 0 12 0z"/>
                   </svg>
                   View source
                </a>
            </div>
        `;
        toc.appendChild(sourceLink);
    }
});
</script>
```
