---
title: "Data Science for Public Policy"
subtitle: "Final Project"
author: 
  - name: "Troy Cheng\u00A0   Ziqiao Shan\u00A0  Minji Kang\u00A0 & \u00A0Sarah Krause"
    email: yc1317@georgetown.edu
    affiliation: Georgetown University
    corresponding: true
df-print: kable
title-block-banner: "#0a0e1a"
title-block-banner-color: "#4DB8FF"
execute:
  warning: false
date: 2025-03-25
date-modified: last-modified
format:
  html:
    embed-resources: true
    toc: true                 
    toc-title: "Contents"     
    toc-location: right       
    number-sections: true     
    smooth-scroll: true       
    css: trycstyle.css 
    code-overflow: wrap
include-in-header:
  text: |
    <link rel="shortcut icon" href="assets/gu.ico" type="image/x-icon">           
highlight-style: nord       
engine: knitr
---

# Exploratory Data Analysis of DC Crime Incidents

## Introduction

This analysis explores crime incident data in Washington DC from 2008 to 2025, focusing on spatial and temporal patterns. We'll examine trends in crime types, geographic distribution, and changes over time to provide insights into crime patterns across the city.

## Load Packages

```{r}
# Load required packages
# install.packages(c("leaflet", "scales", "viridis", "leaflet.extras", "DT", "plotly"))

library(tidyverse)
library(sf)
library(ggmap)
library(leaflet)
library(leaflet.extras) # For heat maps and additional leaflet functionality
library(lubridate)      # For date manipulation
library(scales)         # For better axis formatting
library(viridis)        # For color palettes
library(gridExtra)      # For arranging multiple plots
library(DT)             # For interactive tables
library(plotly)         # For interactive plots
```

## Load and Prepare Crime Data

```{r}
# Set the data directory path
data_dir <- "data/DC/crime_incidents/"

# Load crime data for multiple years (2021-2023 as examples)
crime_2021 <- read_csv(paste0(data_dir, "Crime_Incidents_in_2021.csv"))
crime_2022 <- read_csv(paste0(data_dir, "Crime_Incidents_in_2022.csv"))
crime_2023 <- read_csv(paste0(data_dir, "Crime_Incidents_in_2023.csv"))

# Function to clean crime data
clean_crime_data <- function(data) {
  data %>%
    # Convert date columns to proper date format
    mutate(
      REPORT_DAT = as.Date(str_extract(REPORT_DAT, "\\d{4}/\\d{2}/\\d{2}")),
      START_DATE = as.Date(str_extract(START_DATE, "\\d{4}/\\d{2}/\\d{2}")),
      END_DATE = as.Date(str_extract(END_DATE, "\\d{4}/\\d{2}/\\d{2}"))
    ) %>%
    # Remove rows with missing coordinates
    filter(!is.na(LATITUDE) & !is.na(LONGITUDE))
}

# Clean each dataset
crime_2021_clean <- clean_crime_data(crime_2021)
crime_2022_clean <- clean_crime_data(crime_2022)
crime_2023_clean <- clean_crime_data(crime_2023)

# Combine datasets and add year column
crime_2021_clean <- crime_2021_clean %>% mutate(YEAR = 2021)
crime_2022_clean <- crime_2022_clean %>% mutate(YEAR = 2022)
crime_2023_clean <- crime_2023_clean %>% mutate(YEAR = 2023)

# Combine all years
all_crime <- bind_rows(crime_2021_clean, crime_2022_clean, crime_2023_clean)

# Display the structure of the combined dataset
glimpse(all_crime)
```

## Basic Crime Statistics

```{r}
# Count of incidents by year
all_crime %>%
  count(YEAR) %>%
  knitr::kable(caption = "Number of Crime Incidents by Year")

# Top crime types across all years
all_crime %>%
  count(OFFENSE, sort = TRUE) %>%
  head(10) %>%
  knitr::kable(caption = "Top 10 Crime Types (2021-2023)")

# Crime methods across all years
all_crime %>%
  count(METHOD, sort = TRUE) %>%
  head(10) %>%
  knitr::kable(caption = "Top 10 Crime Methods (2021-2023)")

# Crime by ward
all_crime %>%
  filter(!is.na(WARD) & WARD != "") %>%
  count(WARD, sort = TRUE) %>%
  knitr::kable(caption = "Crime Incidents by Ward (2021-2023)")

# Crime by shift
all_crime %>%
  count(SHIFT, sort = TRUE) %>%
  knitr::kable(caption = "Crime Incidents by Shift (2021-2023)")
```

## Temporal Analysis

```{r, fig.width=10, fig.height=6}
# Crime by month across years
all_crime %>%
  mutate(month = month(REPORT_DAT, label = TRUE)) %>%
  count(YEAR, month) %>%
  ggplot(aes(x = month, y = n, fill = factor(YEAR))) +
  geom_col(position = "dodge") +
  labs(title = "Crime Incidents by Month and Year",
       x = "Month",
       y = "Number of Incidents",
       fill = "Year") +
  theme_minimal() +
  scale_fill_viridis_d()

# Crime by day of week
all_crime %>%
  mutate(day_of_week = wday(REPORT_DAT, label = TRUE)) %>%
  count(day_of_week) %>%
  ggplot(aes(x = day_of_week, y = n, fill = day_of_week)) +
  geom_col() +
  labs(title = "Crime Incidents by Day of Week (2021-2023)",
       x = "Day of Week",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_viridis_d() +
  theme(legend.position = "none")

# Crime by time of day (shift)
all_crime %>%
  count(SHIFT) %>%
  ggplot(aes(x = SHIFT, y = n, fill = SHIFT)) +
  geom_col() +
  labs(title = "Crime Incidents by Time of Day (2021-2023)",
       x = "Shift",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_viridis_d() +
  theme(legend.position = "none")
```

## Crime Type Analysis

```{r, fig.width=10, fig.height=8}
# Top crime types by year
all_crime %>%
  count(YEAR, OFFENSE) %>%
  group_by(YEAR) %>%
  top_n(10, n) %>%
  ungroup() %>%
  ggplot(aes(x = reorder(OFFENSE, n), y = n, fill = factor(YEAR))) +
  geom_col() +
  facet_wrap(~YEAR, scales = "free_y") +
  coord_flip() +
  labs(title = "Top 10 Crime Types by Year",
       x = "Crime Type",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_viridis_d() +
  theme(legend.position = "none")

# Crime methods by year
all_crime %>%
  filter(METHOD != "OTHERS") %>%  # Exclude "OTHERS" to focus on specific methods
  count(YEAR, METHOD) %>%
  group_by(YEAR) %>%
  top_n(5, n) %>%
  ungroup() %>%
  ggplot(aes(x = reorder(METHOD, n), y = n, fill = factor(YEAR))) +
  geom_col() +
  facet_wrap(~YEAR, scales = "free_y") +
  coord_flip() +
  labs(title = "Top 5 Crime Methods by Year",
       x = "Method",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_viridis_d() +
  theme(legend.position = "none")
```

## Spatial Analysis

```{r, fig.width=10, fig.height=8}
# Create a basic map of crime incidents for 2023
leaflet(crime_2023_clean) %>%
  addTiles() %>%  # Add default OpenStreetMap tiles
  addCircleMarkers(
    ~LONGITUDE, ~LATITUDE,
    radius = 3,
    color = "red",
    stroke = FALSE,
    fillOpacity = 0.5,
    popup = ~paste("Offense:", OFFENSE, "<br>Date:", REPORT_DAT, "<br>Block:", BLOCK)
  ) %>%
  addLegend("bottomright", 
            colors = "red",
            labels = "Crime Incidents (2023)",
            opacity = 0.7)

# Create a heat map of crime incidents for 2023
leaflet(crime_2023_clean) %>%
  addTiles() %>%  # Add default OpenStreetMap tiles
  addHeatmap(
    lng = ~LONGITUDE, 
    lat = ~LATITUDE,
    intensity = 0.5,
    radius = 10,
    blur = 15
  ) %>%
  addLegend("bottomright", 
            colors = c("yellow", "red"),
            labels = c("Low Density", "High Density"),
            title = "Crime Density (2023)",
            opacity = 0.7)
```

## Load and Analyze Shapefile Data

```{r}
# Load shapefile data for 2025
crime_2025_sf <- st_read(paste0(data_dir, "shp/Crime_Incidents_in_2025/Crime_Incidents_in_2025.shp"))

# Display the structure of the shapefile data
glimpse(crime_2025_sf)

# Display the first few rows of the shapefile data
head(crime_2025_sf) %>% knitr::kable()

# Plot the spatial distribution of crime incidents from the shapefile
plot(st_geometry(crime_2025_sf), main = "Spatial Distribution of Crime Incidents (2025)")
```

## Combine CSV and Shapefile Data

```{r}
# Load the combined CSV data (2008-2025)
try({
  all_years_csv <- read_csv(paste0(data_dir, "Crime_Incidents_in_2008 - 2025.csv"))
  
  # Clean the combined CSV data
  all_years_clean <- all_years_csv %>%
    # Convert date columns to proper date format
    mutate(
      REPORT_DAT = as.Date(str_extract(REPORT_DAT, "\\d{4}/\\d{2}/\\d{2}")),
      START_DATE = as.Date(str_extract(START_DATE, "\\d{4}/\\d{2}/\\d{2}")),
      END_DATE = as.Date(str_extract(END_DATE, "\\d{4}/\\d{2}/\\d{2}"))
    ) %>%
    # Remove rows with missing coordinates
    filter(!is.na(LATITUDE) & !is.na(LONGITUDE))
  
  # Convert CSV data to sf object
  all_years_sf <- all_years_clean %>%
    st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4326)
  
  # Combine with shapefile data
  combined_sf <- rbind(all_years_sf, crime_2025_sf)
  
  # Plot the combined data
  plot(st_geometry(combined_sf), main = "Combined Crime Incidents (2008-2025)")
}, silent = TRUE)
```

## Crime Trends Over Time

```{r, fig.width=10, fig.height=6}
# Extract year from report date for all crime data
all_crime <- all_crime %>%
  mutate(year = year(REPORT_DAT))

# Count incidents by year
yearly_counts <- all_crime %>%
  count(year)

# Plot crime trends over time
ggplot(yearly_counts, aes(x = year, y = n)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "blue", size = 3) +
  labs(title = "Crime Incidents by Year",
       x = "Year",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_x_continuous(breaks = unique(yearly_counts$year))

# Count incidents by year and offense type (top 5)
top_offenses <- all_crime %>%
  count(OFFENSE) %>%
  top_n(5, n) %>%
  pull(OFFENSE)

offense_trends <- all_crime %>%
  filter(OFFENSE %in% top_offenses) %>%
  count(year, OFFENSE)

# Plot trends for top offense types
ggplot(offense_trends, aes(x = year, y = n, color = OFFENSE, group = OFFENSE)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Trends for Top 5 Crime Types",
       x = "Year",
       y = "Number of Incidents",
       color = "Offense Type") +
  theme_minimal() +
  scale_x_continuous(breaks = unique(offense_trends$year)) +
  scale_color_viridis_d()
```

## Ward-Level Analysis

```{r, fig.width=10, fig.height=8}
# Crime incidents by ward and year
ward_year_counts <- all_crime %>%
  filter(!is.na(WARD) & WARD != "") %>%
  count(WARD, YEAR)

# Plot crime by ward and year
ggplot(ward_year_counts, aes(x = WARD, y = n, fill = factor(YEAR))) +
  geom_col(position = "dodge") +
  labs(title = "Crime Incidents by Ward and Year",
       x = "Ward",
       y = "Number of Incidents",
       fill = "Year") +
  theme_minimal() +
  scale_fill_viridis_d()

# Analyze top crime types by ward
ward_crime_types <- all_crime %>%
  filter(!is.na(WARD) & WARD != "") %>%
  count(WARD, OFFENSE) %>%
  group_by(WARD) %>%
  top_n(3, n) %>%
  ungroup()

# Create a faceted bar chart of top crime types by ward
ggplot(ward_crime_types, aes(x = reorder(OFFENSE, n), y = n, fill = OFFENSE)) +
  geom_col() +
  facet_wrap(~WARD, scales = "free_y") +
  coord_flip() +
  labs(title = "Top 3 Crime Types by Ward (2021-2023)",
       x = "Crime Type",
       y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_viridis_d() +
  theme(legend.position = "none")
```

## Conclusion

This exploratory data analysis of crime incidents in Washington DC from 2021 to 2023 reveals several important patterns:

1. **Temporal Patterns**: Crime incidents show distinct patterns by time of day, day of week, and season, with certain periods experiencing higher rates of specific crimes.

2. **Spatial Distribution**: Heat maps highlight crime hotspots across the city, with different crime types showing unique spatial patterns.

3. **Crime Trends**: The analysis of multi-year data shows changing patterns in crime types and methods over time, with some offenses increasing while others decrease.

4. **Ward-Level Variations**: Different wards experience different types and frequencies of crime, suggesting the need for targeted prevention strategies.

Further analysis could explore the relationship between crime patterns and socioeconomic factors, as well as the impact of policy interventions on crime rates in specific areas.


```{=html}
<script>
document.addEventListener("DOMContentLoaded", function() {
    const toc = document.getElementById("TOC");
    if (toc) {
        const sourceLink = document.createElement("div");
        sourceLink.innerHTML = `
            <div class="toc-source">
                <a href="https://github.com/troy-yu-cheng/final-project" 
                   target="_blank" 
                   class="github-button">
                   <svg xmlns="http://www.w3.org/2000/svg" 
                        viewBox="0 0 24 24" 
                        width="16" 
                        height="16" 
                        fill="currentColor"
                        style="vertical-align: middle; margin-right: 5px;">
                     <path d="M12 0C5.373 0 0 5.373 0 12c0 5.303 3.438 9.8 8.207 11.387.6.113.82-.26.82-.577v-2.157c-3.338.726-4.033-1.416-4.033-1.416-.546-1.386-1.332-1.756-1.332-1.756-1.09-.745.083-.73.083-.73 1.205.084 1.84 1.237 1.84 1.237 1.07 1.832 2.807 1.303 3.492.996.108-.774.418-1.303.76-1.602-2.665-.3-5.466-1.332-5.466-5.93 0-1.311.468-2.382 1.237-3.222-.124-.302-.536-1.52.118-3.163 0 0 1.008-.322 3.3 1.23a11.516 11.516 0 0 1 3.002-.403 11.486 11.486 0 0 1 3.002.403c2.292-1.552 3.3-1.23 3.3-1.23.654 1.644.242 2.861.118 3.163.77.84 1.236 1.911 1.236 3.222 0 4.61-2.807 5.627-5.48 5.922.43.372.812 1.103.812 2.222v3.293c0 .321.218.694.825.576C20.565 21.796 24 17.3 24 12 24 5.373 18.627 0 12 0z"/>
                   </svg>
                   View source
                </a>
            </div>
        `;
        toc.appendChild(sourceLink);
    }
});
</script>
